---
title: 'EDS 241: Assignment 1'
author: 'Javier Patrón'
date: "`r Sys.Date()`"
output:
  pdf_document:
    toc: false
editor_options: 
  markdown: 
    wrap: 72
---

## Question 1:

Application of estimators based on the "treatment ignorability"
assumption. The goal is to estimate the causal effect of maternal
smoking during pregnancy on infant birth weight using the treatment
ignorability assumptions (Lecture 6 & 7).

The data are taken from the National Natality Detail Files, and the
extract "SMOKING_EDS241.csv"' is a random sample of all births in
Pennsylvania during 1989-1991. Each observation is a mother-infant pair.
The key variables are:

The outcome and treatment variables are:

-   birthwgt = birth weight of infant in grams

-   tobacco = indicator for maternal smoking

The control variables are: mage (mother's age), meduc (mother's
education), mblack (=1 if mother black), alcohol (=1 if consumed alcohol
during pregnancy), first (=1 if first child), diabete (=1 if mother
diabetic), anemia (=1 if mother anemic)

```{r}
library(tidyverse)
library(stargazer)
library(estimatr)
library(corrplot)
library(tidymodels)

```


```{r}
df_smoking <- read_csv(here::here("SMOKING_EDS241.csv"))
```
What is the unadjusted mean difference in birth weight of infants with smoking and nonsmoking mothers? 

```{r}
mean_weight_smk_nsmk <- df_smoking |> 
  group_by(tobacco) |> 
  summarise(mean_weight = mean(birthwgt))

round(mean_weight_smk_nsmk$mean_weight[1] - mean_weight_smk_nsmk$mean_weight[2],2 )
```
The unadjusted mean difference in birth weight of infants with smoking and nonsmoking mothers is 244.54 



Under what assumption does this correspond to the average treatment effect of maternal smoking during pregnancy on infant birth weight? Provide some simple empirical evidence for or against this assumption.
```{r}
df_smoking |>
  mutate(tobacco = case_when(tobacco == 0 ~ "Non Smoker",
                             tobacco == 1 ~ "Smoker")) |>
  ggplot(aes(
    x = as.factor(tobacco),
    y = meduc,
    fill = tobacco
  )) +
  geom_boxplot(alpha = 0.7) +
  labs(x = "Tobacco",
       y = "Birth Weight (grams)",
       title = "Difference between smoking and non-smoking mothers in Birth weight ")
```


Another way of seeing this is by the lm() function

```{r}
lm(data = smoke, birthwgt ~ tobacco)
```

Thanks to the results above with lm() we can see the same result as
above. The "slope" tell us that the difference between weight babies
between smokers and nonsmokers is -244.5 grams. For every "unit" which
is YES or NO, a baby on average will be 244.5 grams lighter **omitting
all of the other variables**.

This assumption is under the fact that the 244.5 grams is the decrease
on weight baby in case the mother smokes on pregnancy. But is is **not
the correct assumption**. There are other factors that can have a
significant effect on the birth weight, as you can see in this
correlation matrix

```{r, echo = FALSE}

# Obtain correlation matrix
matrix <- cor(smoke)

# Make a correlation plot between the variables
corrplot(matrix, 
         method = "color", 
         shade.col = NA, 
         tl.col = "gray40", 
         tl.srt = 45, 
         addCoef.col = "gray20", 
         cl.pos = "n", 
         order = "hclust",
         col = colorRampPalette(c("red", "white", "deepskyblue4"))(100))

```

## Question (b)

#### Assume that maternal smoking is randomly assigned conditional on the observable covariates listed above. Estimate the effect of maternal smoking on birth weight using an OLS regression with linear controls for the covariates. Report the estimated coefficient on tobacco and its standard error.

```{r}
# Question (b)
# Fit the linear regression model with all the variables
all <- lm(data = smoke, birthwgt ~ .)

# Create a pretty table of the results
x <- starprep(all, 
             stat = c("std.error"), 
             se_type = "HC1", 
             alpha = 0.05)

stargazer(all, 
           se = x, 
           type="text")

```

As you can see in the table above the tobacco impact in birth weight has
moved from 244.5 to 228.073. This means that tobacco is not the only
variable that is affecting a possible decrease in weight for newborns.
There are other factors and variables affecting the weight birth of
infants.

Additionally thanks to the stargazer function we can see the
heteroskedasticity error for tobacco is 4.277. Meaning that the related
difference of 228.073 grams that smoker mothers have on their child
birth weight will have a has a standard error of 4.277 grams on average.
In other words, the heteroskedasticity will have a 4.277 grams of
possible error in this estimation.

(c) Use the exact matching estimator to estimate the effect of maternal
    smoking on birth weight. For simplicity, consider the following
    covariates in your matching estimator: create a 0-1 indicator for
    mother's age (= 1 if mage\> = 34), and a 0-1 indicator for mother's
    education (1 if meduc\> = 16), mother's race (mblack), and alcohol
    consumption indicator (alcohol). These 4 covariates will create 2 \*
    2 *\** 2 \* 2 = 16 cells.

```{r}
# Question (c)
#Create the new column with the code
smoke_new <- smoke |> 
  mutate(mage = case_when(mage > 34 ~ 1,
                          TRUE ~ 0)) |> 
  mutate(meduc = case_when(meduc > 16 ~ 1,
                           TRUE ~0)) |> 
  mutate(code = as.factor(paste0(mage, meduc, mblack,alcohol)))

class(smoke_new$code)


# MULTIVARIATE MATCHING AS REGRESSION ESTIMATOR
reg_ate <- lm(data = smoke_new, birthwgt ~ tobacco + code)

se_models = starprep(reg_ate, stat = c("std.error"), se_type = "HC2", alpha = 0.05)
stargazer(reg_ate, se = se_models, type="text")


ggplot(smoke_new, aes(x=code, y = birthwgt, fill = code)) + 
  geom_boxplot(alpha = 0.5) +
  labs(x = "Covariets",
       y = "Birth Weight (g)",
       title = "")


```

Report the estimated average treatment effect of smoking on birthweight
using the exact matching estimator and its linear regression analogue.
